---
title: "Figure 2"
---

## Overview

Figure 2 evaluates the performance of `geokit` for representative GEO data reuse
tasks. The benchmark focuses on two common operations: parsing complete GEO SOFT
records and extracting metadata from a disease-specific GEO collection. The
comparison includes the Rust-backed implementation in `geokit`, the earlier
data.table-backed implementation, and `GEOquery` where applicable. 

The SOFT parsing benchmark uses accession `GSE67705` and compares elapsed time
and memory allocation. The metadata benchmark uses GEO Series records returned
from a real-time NCBI E-utils search for human diabetes datasets and compares
elapsed time for metadata parsing. 

To reduce the influence of network variation, GEO files are first downloaded to `rawdata/Figure 2`. The benchmark therefore mainly reflects parsing and metadata extraction from locally cached files.

## Required packages

The following packages are required.

```{r}
#| code-fold: true
#| output: false
#| eval: false
pak::pak("bench")
pak::pak("ggplot2")
pak::pak("GEOquery")
pak::pak("WangLabCSU/geokit")
pak::pak("WangLabCSU/geokit@49717ddc616b684e916e165b4eb508ca492f9ccf") # install rgeo
````

## Setup
```{r}
#| code-fold: true
#| output: false
if (!dir.exists("rawdata/Figure 2")) {
    dir.create("rawdata/Figure 2", recursive = TRUE)
}
if (!dir.exists("Figure 2")) {
    dir.create("Figure 2")
}

library(GEOquery)
library(rgeo) # data.table-backed implementation
library(geokit) # Rust-backed implementation
library(ggplot2)
```

## GEO search and metadata collection

This section constructs a disease-specific GEO collection using a real-time
E-utils query. The query retrieves human GEO Series records associated with
diabetes. These records are then used as input for the metadata benchmark.
Before benchmarking, `geo_meta()` is run once to download and cache the required
GEO files.

```{r}
diabetes_records <- geo_search(
    "diabetes[ALL] AND Homo sapiens[ORGN] AND GSE[ETYP]"
)
diabetes_meta <- geo_meta(
    diabetes_records[["Series Accession"]],
    handle_opts = list(connecttimeout = 600L),
    odir = "rawdata/Figure 2"
)
# Download data, GEOquery use different naming convention
getGEO( # GSE67705.soft.gz
    "GSE67705",
    GSEMatrix = FALSE,
    destdir = "rawdata/Figure 2"
)
```

## Benchmark 1: GEO SOFT parsing

The first benchmark compares complete GEO SOFT parsing for `GSE67705`. This task
evaluates the ability of each implementation to parse a full GEO Series SOFT
record. Both timing and memory allocation are recorded for this benchmark.

```{r}
#| code-fold: false
#| eval: false
comparative_soft <- bench::mark(
    `geokit(rust)` = geo_soft(
        "GSE67705",
        odir = "rawdata/Figure 2"
    ),
    `geokit(data.table)` = get_geo( # GSE67705_family.soft.gz
        "GSE67705",
        gse_matrix = FALSE,
        dest_dir = "rawdata/Figure 2"
    ),
    GEOquery = getGEO( # GSE67705.soft.gz
        "GSE67705",
        GSEMatrix = FALSE,
        destdir = "rawdata/Figure 2"
    ),
    max_iterations = 1L,
    check = FALSE,
    memory = TRUE
)
saveRDS(comparative_soft, "Figure 2/comparative_soft.rds")
```

## Benchmark 2: GEO metadata retrieval

The second benchmark compares metadata retrieval for the diabetes-related GEO
Series accessions identified above. This benchmark focuses on elapsed time only.

```{r}
#| code-fold: false
#| eval: false
comparative_meta <- bench::mark(
    `geokit(rust)` = geo_meta(
        diabetes_records[["Series Accession"]],
        odir = "rawdata/Figure 2"
    ),
    `geokit(data.table)` = get_geo_meta(
        diabetes_records[["Series Accession"]],
        dest_dir = "rawdata/Figure 2"
    ),
    max_iterations = 1L,
    check = FALSE,
    memory = FALSE
)
saveRDS(comparative_meta, "Figure 2/comparative_meta.rds")
```

## Load benchmark results

```{r}
soft_rds <- "Figure 2/comparative_soft.rds"
meta_rds <- "Figure 2/comparative_meta.rds"
if (!file.exists(soft_rds)) {
    stop("Missing benchmark result: ", soft_rds)
}
if (!file.exists(meta_rds)) {
    stop("Missing benchmark result: ", meta_rds)
}
file.info("rawdata/Figure 2/GSE67705_family.soft.gz")$size / 1024^3
file.info("rawdata/Figure 2/GSE67705.soft.gz")$size / 1024^3
comparative_soft <- readRDS(soft_rds)
comparative_meta <- readRDS(meta_rds)
```

## SOFT parsing time

```{r}
#| label: fig-soft-time
#| fig-cap: "Elapsed time for GEO SOFT parsing of GSE67705."
#| fig-width: 6
#| fig-height: 3.5
A <- comparative_soft |>
    dplyr::mutate(
        id = names(expression),
        id = factor(id, unique(id))
    ) |>
    ggplot(aes(x = id, y = total_time)) +
    geom_col(aes(fill = id, color = after_scale(fill)), width = 0.7) +
    geom_text(
        aes(label = sprintf("%ds", as.integer(total_time))),
        vjust = -0.5
    ) +
    scale_fill_brewer(palette = "Set2", guide = "none") +
    labs(x = NULL, y = "Elapsed time (seconds)") +
    theme_classic(base_size = 12)
ggsave("Figure 2/A.pdf", plot = A, height = 5, width = 6)
```

## SOFT parsing memory allocation

```{r}
#| label: fig-soft-memory
#| fig-cap: "Memory allocation for GEO SOFT parsing of GSE67705."
#| fig-width: 6
#| fig-height: 3.5
B <- comparative_soft |>
    dplyr::mutate(
        id = names(expression),
        id = factor(id, unique(id))
    ) |>
    ggplot(aes(x = id, y = mem_alloc / 1024^3)) +
    geom_col(aes(fill = id, color = after_scale(fill)), width = 0.7) +
    geom_text(
        aes(label = sprintf("%.2fGB", mem_alloc / 1024^3)),
        vjust = -0.5
    ) +
    scale_fill_brewer(palette = "Set2", guide = "none") +
    labs(x = NULL, y = "Memory allocation (GB)") +
    theme_classic(base_size = 12)
ggsave("Figure 2/B.pdf", plot = B, height = 5, width = 6)
```

## Metadata retrieval time

```{r}
#| label: fig-meta-time
#| fig-cap: "Elapsed time for metadata parsing from diabetes-related human GEO Series records."
#| fig-width: 6
#| fig-height: 3.5
C <- comparative_meta |>
    dplyr::mutate(
        id = names(expression),
        id = factor(id, unique(id))
    ) |>
    ggplot(aes(x = id, y = total_time / 60)) +
    geom_col(aes(fill = id, color = after_scale(fill)), width = 0.7) +
    geom_text(
        aes(label = sprintf("%.2fmin", total_time / 60)),
        vjust = -0.5
    ) +
    scale_fill_brewer(palette = "Set2", guide = "none") +
    labs(x = NULL, y = "Elapsed time (minutes)") +
    theme_classic(base_size = 12)
ggsave("Figure 2/C.pdf", plot = C, height = 5, width = 4)
```
